Reconstructing the neutron-star equation of state with gravitational-wave detectors from a realistic population of inspiralling binary neutron stars

### 2014-09-16 paper.tex
Read through Ben's changes and answered questions

\documentclass[twocolumn,prd,amssymb,aps,nofootinbib,showpacs,epsf]{revtex4}
\usepackage[usenames]{color}
\usepackage{ulem}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{epstopdf}

@@

For all versions of the PN waveform, we cut off the inspiral at the GW frequency corresponding to the Schwarzschild innermost stable circular orbit (ISCO) $f_{\rm ISCO}=(6^{3/2}\pi G M/c^3)$.  However, for large NS radii corresponding to a stiff EOS, NSs can merge before reaching $f_{\rm ISCO}$. For simplicity, we also choose to examine nonspinning BNS systems, although it has been shown that not including spin parameters can noticeably bias parameter estimation even for systems with relatively small spin magnitudes \cite{Favata2014}. A parallel investigation which studies how the choice of high-frequency cutoffs and spin can affect the estimation of tidal parameters is nearing completion \cite{Nikhef_paper_2} \red{[[Get reference]]}.

@@

For each piece of a piecewise polytrope, the pressure $p$ in the rest-mass density interval $\rho_{i-1} < \rho < \rho_i$ is defined by
\begin{equation}
p(\rho) = K_i \rho^{\Gamma_i},
\end{equation}
where $\Gamma_i$ is the adiabatic index and the constant $K_i$ is chosen such that $p$ is continuous at the boundary $\rho_{i-1}$. For the core of the star, we use 3 polytropes with adiabatic indices $\Gamma_1$, $\Gamma_2$, and $\Gamma_3$, separated by the fixed dividing densities $\rho_1 = 10^{14.7}$~g/cm$^3$ and $\rho_2 = 10^{15}$~g/cm$^3$. These fixed dividing densities were chosen to minimize the least-squares error between the piecewise polytrope fit and a set of 34 tabulated theoretical EOSs~\cite{ReadLackey2009}. In addition to the three adiabatic indices, we also require an additional parameter to determine the overall pressure scaling which we choose to be $p_1 = p(\rho_1)$, the pressure at the first dividing density. We will therefore use as our four EOS parameters $\vec E = \{\log(p_1), \Gamma_1, \Gamma_2, \Gamma_3\}$. \les{($E$ was used as energy in Section II.  We should change one or the other, I think)}{} \ben{[[I think it's clear from context and the fact that the parameters $\vec E$ are always a vector. Was this not true?]]}{}

@@

Given the GW model and EOS fit described in Sections~\ref{sec:waveform} and~\ref{sec:eos}, we will describe here a method for estimating the EOS parameters from a set of GW observations. We use Bayes' theorem to write the posterior density function (PDF) for the universal EOS parameters $\vec E = \{\log(p_1), \Gamma_1, \Gamma_2, \Gamma_3\}$ that are common to all NSs \les{(The referee made me provide a citation for this type of statement)}{}\ben{[[What citation did you use?]]}{} and the waveform parameters $\vec\theta = \{d_L, \alpha, \delta, \psi, \iota, t_c, \phi_c, m_1, m_2, \tilde\Lambda, \delta\tilde\Lambda\}$ that are unique to each of the $n$ BNS systems as
\begin{equation}
\begin{split}
p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I})
&= p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I}) \\
& \times\frac{ p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I}) }{ p(D | \mathcal{H},\mathcal{I}) }.
\end{split}
\end{equation}
The data $D = \vec d_1,\dots, \vec d_n$ is composed of the data streams $\vec d_i(t)$ from the GW detector network for each of the $n$ inspiral events. $\mathcal{H}$ represents the EOS and waveform models, and $\mathcal{I}$ represents the background information for the parameters. The quantity $p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I})$ is the prior probability density for the EOS and waveform parameters, $p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I})$ is the likelihood, and the normalization constant $p(D | \mathcal{H},\mathcal{I})$, which we will not need to calculate here, is the evidence. The marginalized PDF for the EOS parameters is then
\begin{equation}
\label{eq:margEOS}
p(\vec E | D,\mathcal{H},\mathcal{I}) = \int d\vec\theta_1 \dots d\vec\theta_n p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I}).
\end{equation}
This $11n$-dimensional integral is not easily computed, so we will decompose it into blocks that can be individually computed. First, we decompose the waveform parameters into extrinsic \red{[[should you call these nuisance parameters instead?]]} parameters $\vec\theta_{\rm ex} = \{D_L, \alpha, \delta, \psi, \iota, t_c, \phi_c, \delta\tilde\Lambda\}$\footnote{$\delta\tilde\Lambda$ is an intrinsic parameter that in principle provides additional EOS information. However, it is unmeasurable with aLIGO~\cite{WadeCreightonOchsner2014}, so we treat it as a nuisance parameter and group it with the extrinsic parameters when we marginalize over them.} that depend on the observer and intrinsic parameters $\vec\theta_{\rm in} = \{m_1, m_2, \tilde\Lambda\}$ that are relevant for measuring the EOS such that $\vec\theta = \{\vec\theta_{\rm ex}, \vec\theta_{\rm in}\}$.

@@

In this Section we characterize the ability of the aLIGO--aVirgo network to measure the EOS from a population of BNS inspiral events. For the two aLIGO detectors, we use the zero detuned-high laser power (broadband) PSD~\cite{LIGOnoise} which represents the design sensitivity that may be achieved for the aLIGO detectors by 2019~\cite{AasiAbadieAbbott2013}. For aVIRGO, which has higher high-frequency noise, we use the PSD fit from Ref.~\cite{ManzottiDietz2012} that may be achieved by 2021~\cite{AasiAbadieAbbott2013}. \ben{[[Why is the aVirgo PSD a few times higher between 500--1000Hz than the aLIGO one in the reference (Fig. 1)? I've seen noise curves where the high frequency is more comparable. Do you know of a reference that describes the different aVirgo configurations and noise curves?]]}{}

@@

We simulated a year of GW data using the ``realistic'' event rate above. Specifically, we calculated the event rate in a volume large enough to contain all detectable BNS events. Because inspirals are a Poisson process, we sampled the actual number of events in a year from a Poisson distribution with this rate. We then sampled the locations of these events uniformly in the volume, the orientations uniformly on a unit sphere, and the individual NS masses uniformly in $[1.2M_\odot, 1.4M_\odot]$. Of these systems $\sim$120 had a network SNR $\ge$ 8 and $\sim$30 had a network SNR $\ge$ 12. We performed parameter estimation for the 20 loudest sources with network SNR that ranged from 63.7 down to 13.6, integrating between GW frequencies \ben{$f_{\rm low}=10?30?$~Hz}{} and $f_{\rm ISCO}$.

>>>>

\documentclass[twocolumn,prd,amssymb,aps,nofootinbib,showpacs,epsf]{revtex4}
\usepackage[usenames]{color}
\usepackage{ulem}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{hyperref}

@@

For all versions of the PN waveform, we cut off the inspiral at the GW frequency corresponding to the Schwarzschild innermost stable circular orbit (ISCO) \les{$f_{\rm ISCO}=c^3/(6^{3/2}\pi G M)$ }{$f_{\rm ISCO}=(6^{3/2}\pi G M/c^3)$}.  However, for large NS radii corresponding to a stiff EOS, NSs can merge before reaching $f_{\rm ISCO}$. For simplicity, we also choose to examine nonspinning BNS systems, although it has been shown that not including spin parameters can noticeably bias parameter estimation even for systems with relatively small spin magnitudes \cite{Favata2014}. A parallel investigation which studies how the choice of high-frequency cutoffs and spin can affect the estimation of tidal parameters is nearing completion \cite{Nikhef_paper_2} \red{[[Get reference]]}.

@@

For each piece of a piecewise polytrope, the pressure $p$ in the rest-mass density interval $\rho_{i-1} < \rho < \rho_i$ is defined by
\begin{equation}
p(\rho) = K_i \rho^{\Gamma_i},
\end{equation}
where $\Gamma_i$ is the adiabatic index and the constant $K_i$ is chosen such that $p$ is continuous at the boundary $\rho_{i-1}$. For the core of the star, we use 3 polytropes with adiabatic indices $\Gamma_1$, $\Gamma_2$, and $\Gamma_3$, separated by the fixed dividing densities $\rho_1 = 10^{14.7}$~g/cm$^3$ and $\rho_2 = 10^{15}$~g/cm$^3$. These fixed dividing densities were chosen to minimize the least-squares error between the piecewise polytrope fit and a set of 34 tabulated theoretical EOSs~\cite{ReadLackey2009}. In addition to the three adiabatic indices, we also require an additional parameter to determine the overall pressure scaling which we choose to be $p_1 = p(\rho_1)$, the pressure at the first dividing density. We will therefore use as our four EOS parameters $\vec E = \{\log(p_1), \Gamma_1, \Gamma_2, \Gamma_3\}$. \les{($E$ was used as energy in Section II.  We should change one or the other, I think)}{} \ben{[[I think it's clear from context and the fact that the parameters $\vec E$ are always a vector. Was this not true?]]}{} \les{(Yes, this is probably true.)}{}

@@

Given the GW model and EOS fit described in Sections~\ref{sec:waveform} and~\ref{sec:eos}, we will describe here a method for estimating the EOS parameters from a set of GW observations. We use Bayes' theorem to write the posterior density function (PDF) for the universal EOS parameters $\vec E = \{\log(p_1), \Gamma_1, \Gamma_2, \Gamma_3\}$ that are common to all NSs \les{(The referee made me provide a citation for this type of statement)}{}\ben{[[What citation did you use?]]}{} \les{(``Since all cold NSs have the same EOS [29], spherical NSs with the same mass will also have the same value for $\Lambda$'', where [29] was N. K. Glendenning, Compact Stars: Nuclear Physics, Particle Physics, and General Relativity (Springer- Verlag, NewYork, 2000), 2nd ed.)}{} and the waveform parameters $\vec\theta = \{d_L, \alpha, \delta, \psi, \iota, t_c, \phi_c, m_1, m_2, \tilde\Lambda, \delta\tilde\Lambda\}$ that are unique to each of the $n$ BNS systems as
\begin{equation}
\begin{split}
p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I})
&= p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I}) \\
& \times\frac{ p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I}) }{ p(D | \mathcal{H},\mathcal{I}) }.
\end{split}
\end{equation}
The data $D = \vec d_1,\dots, \vec d_n$ is composed of the data streams $\vec d_i(t)$ from the GW detector network for each of the $n$ inspiral events. $\mathcal{H}$ represents the EOS and waveform models, and $\mathcal{I}$ represents the background information for the parameters. The quantity $p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I})$ is the prior probability density for the EOS and waveform parameters, $p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I})$ is the likelihood, and the normalization constant $p(D | \mathcal{H},\mathcal{I})$, which we will not need to calculate here, is the evidence. The marginalized PDF for the EOS parameters is then
\begin{equation}
\label{eq:margEOS}
p(\vec E | D,\mathcal{H},\mathcal{I}) = \int d\vec\theta_1 \dots d\vec\theta_n p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I}).
\end{equation}
This $11n$-dimensional integral is not easily computed, so we will decompose it into blocks that can be individually computed. First, we decompose the waveform parameters into extrinsic \red{[[should you call these nuisance parameters instead?]]} parameters $\vec\theta_{\rm ex} = \{D_L, \alpha, \delta, \psi, \iota, t_c, \phi_c, \delta\tilde\Lambda\}$\footnote{$\delta\tilde\Lambda$ is an intrinsic parameter that in principle provides additional EOS information. However, it is unmeasurable with aLIGO~\cite{WadeCreightonOchsner2014}, so we treat it as a nuisance parameter and group it with the extrinsic parameters when we marginalize over them.} that depend on the observer and intrinsic parameters $\vec\theta_{\rm in} = \{m_1, m_2, \tilde\Lambda\}$ that are relevant for measuring the EOS such that $\vec\theta = \{\vec\theta_{\rm ex}, \vec\theta_{\rm in}\}$.

@@

In this Section we characterize the ability of the aLIGO--aVirgo network to measure the EOS from a population of BNS inspiral events. For the two aLIGO detectors, we use the zero detuned-high laser power (broadband) PSD~\cite{LIGOnoise} which represents the design sensitivity that may be achieved for the aLIGO detectors by 2019~\cite{AasiAbadieAbbott2013}. For aVIRGO, which has higher high-frequency noise, we use the PSD fit from Ref.~\cite{ManzottiDietz2012} that may be achieved by 2021~\cite{AasiAbadieAbbott2013}. \ben{[[Why is the aVirgo PSD a few times higher between 500--1000Hz than the aLIGO one in the reference (Fig. 1)? I've seen noise curves where the high frequency is more comparable. Do you know of a reference that describes the different aVirgo configurations and noise curves?]]}{} \les{(I don't know of a reference for different aVirgo configurations. Here's a link to the PSDs in LAL: \url{https://www.lsc-group.phys.uwm.edu/daswg/projects/lal/nightly/docs/html/_l_a_l_sim_noise_p_s_d_8c.html}. Maybe this is helpful?)}{}

@@

We simulated a year of GW data using the ``realistic'' event rate above. Specifically, we calculated the event rate in a volume large enough to contain all detectable BNS events. Because inspirals are a Poisson process, we sampled the actual number of events in a year from a Poisson distribution with this rate. We then sampled the locations of these events uniformly in the volume, the orientations uniformly on a unit sphere, and the individual NS masses uniformly in $[1.2M_\odot, 1.4M_\odot]$. Of these systems $\sim$120 had a network SNR $\ge$ 8 and $\sim$30 had a network SNR $\ge$ 12. We performed parameter estimation for the 20 loudest sources with network SNR that ranged from 63.7 down to 13.6, integrating between GW frequencies \ben{$f_{\rm low}=10?30?$~Hz}{} \les{(30 Hz)}{} and $f_{\rm ISCO}$.
